---
title: "3D brain visualization with ggseg3d"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3D brain visualization with ggseg3d}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  eval = TRUE,
  fig.width = 7,
  fig.height = 5
)
library(dplyr)
library(tidyr)
```

ggseg3d renders interactive 3D brain atlases using WebGL.
Version 2.0 rebuilt the package on Three.js, replacing the previous Plotly backend.
This brings faster rendering, a pipe-friendly API, and unified brain atlases that work with both ggseg and ggseg3d.

```{r}
#| label: load-packages
library(ggseg3d)
library(dplyr)
```

## Basic usage

Call `ggseg3d()` without arguments to plot the Desikan-Killiany atlas:

```{r}
#| label: basic
#| fig-cap: "Default Desikan-Killiany atlas rendered with ggseg3d."
ggseg3d()
```

The widget is interactive:
- **Rotate**: click and drag
- **Zoom**: scroll wheel
- **Pan**: right-click and drag

Hover over regions to see their names.

## Camera positions

Use `pan_camera()` to set the viewing angle:

```{r}
#| label: camera-preset
#| fig-cap: "Left lateral view of the Desikan-Killiany atlas."
ggseg3d() |>
  pan_camera("left lateral")
```

Available presets:
- `left lateral`, `right lateral`
- `left medial`, `right medial`
- `left superior`, `right superior`
- `left inferior`, `right inferior`
- `left anterior`, `right anterior`
- `left posterior`, `right posterior`

For custom positions, pass a list:

```{r}
#| label: camera-custom
#| eval: false
ggseg3d() |>
  pan_camera(list(eye = list(x = -200, y = 100, z = 100)))
```

## Plotting your data

Provide data with a `region` or `label` column matching the atlas.
Map it to `colour` for the fill and optionally `text` for hover info:

```{r}
#| label: plot-data
#| fig-cap: "Brain regions coloured by p-value."
some_data <- tibble(
  region = c("precentral", "postcentral", "insula", "superior parietal"),
  p = c(0.01, 0.04, 0.2, 0.5)
)

ggseg3d(.data = some_data, atlas = dk, colour = "p", text = "p") |>
  pan_camera("right lateral")
```

## Custom colour palettes

Pass colours as a vector:

```{r}
#| label: palette-vector
#| eval: false
ggseg3d(
  .data = some_data,
  atlas = dk,
  colour = "p",
  palette = c("forestgreen", "white", "firebrick")
)
```

Or use a named vector to control breakpoints:

```{r}
#| label: palette-named
#| fig-cap: "Custom green-white-red palette with breakpoints at 0, 0.05, and 1."
ggseg3d(
  .data = some_data,
  atlas = dk,
  colour = "p",
  text = "p",
  palette = c("forestgreen" = 0, "white" = .05, "firebrick" = 1)
)
```

This lets the colour bar extend beyond your data range.

## Background colour

Change the background with `set_background()`:

```{r}
#| label: background
#| fig-cap: "Brain atlas with a black background."
ggseg3d() |>
  set_background("black")
```

## Region edges

Add outlines around regions with `set_edges()`:

```{r}
#| label: edges
#| fig-cap: "Left hemisphere with black region boundary edges."
ggseg3d(hemisphere = "left") |>
  set_edges("black", width = 10) |>
  pan_camera("left lateral")
```

### Edge boundaries with edge_by

By default, edges appear between regions with different colours.
Use `edge_by` to define boundaries based on a different column:

```{r}
#| label: edge-by
#| eval: false
some_data <- tibble(
  region = c("precentral", "postcentral", "insula", "superior parietal"),
  p = c(0.2, 0.4, 0.3, 0.5),
  lobe = c("frontal", "parietal", "insula", "parietal")
)

ggseg3d(.data = some_data, atlas = dk, colour = "p", edge_by = "lobe") |>
  set_edges("white", width = 1) |>
  set_background("black")
```

Regions in the same lobe won't have edges between them.

## Subcortical structures

For subcortical atlases, reduce opacity of uncoloured regions and add a glass brain:

```{r}
#| label: subcortical
#| fig-cap: "Subcortical structures with a translucent glass brain overlay."
some_data <- tibble(
  region = c("thalamus", "caudate", "hippocampus"),
  p = c(0.2, 0.5, 0.8)
)

ggseg3d(
  .data = some_data,
  atlas = aseg,
  colour = "p",
  text = "p",
  na_alpha = .5
) |>
  add_glassbrain()
```

The glass brain adds a translucent cortical surface for anatomical context.

## Saving images

Use `snapshot_brain()` to save static images:

```{r}
#| label: snapshot
#| eval: false
ggseg3d() |>
  pan_camera("left lateral") |>
  snapshot_brain("brain_lateral.png")
```

This requires Chrome or Chromium.

### Publication-quality settings

For journal figures, increase resolution:

```{r}
#| label: snapshot-hires
#| eval: false
ggseg3d(.data = some_data, atlas = dk, colour = "p") |>
  pan_camera("left lateral") |>
  set_background("white") |>
  snapshot_brain(
    "figure1_brain.png",
    width = 1200,
    height = 1000,
    zoom = 3,
    delay = 2
  )
```

- **width/height**: Output dimensions in pixels
- **zoom**: Resolution multiplier (2-3 for print)
- **delay**: Seconds to wait for rendering

### Multi-panel figures

Save individual views and combine them:

```{r}
#| label: multi-panel
#| eval: false
library(magick)

base_plot <- ggseg3d(.data = some_data, atlas = dk, colour = "p") |>
  set_background("white")

base_plot |>
  pan_camera("left lateral") |>
  snapshot_brain("left_lat.png")
base_plot |>
  pan_camera("left medial") |>
  snapshot_brain("left_med.png")
base_plot |>
  pan_camera("right lateral") |>
  snapshot_brain("right_lat.png")
base_plot |>
  pan_camera("right medial") |>
  snapshot_brain("right_med.png")

panels <- image_read(c(
  "left_lat.png",
  "left_med.png",
  "right_lat.png",
  "right_med.png"
))
combined <- image_montage(panels, geometry = "600x500", tile = "2x2")
image_write(combined, "figure1_all_views.png")
```

## Legend control

Hide the legend:

```{r}
#| label: legend
#| eval: false
ggseg3d() |>
  set_legend(FALSE)
```

## Widget dimensions

Set explicit width and height:

```{r}
#| label: dimensions
#| eval: false
ggseg3d() |>
  set_dimensions(width = 800, height = 600)
```

## Flat shading

Disable lighting for exact colour reproduction:

```{r}
#| label: flat-shading
#| eval: false
ggseg3d() |>
  set_flat_shading()
```

This is useful when creating atlases or when you need colours to match exactly.

## Orthographic projection

Switch from perspective to orthographic projection:

```{r}
#| label: orthographic
#| eval: false
ggseg3d() |>
  set_orthographic()
```

Regions appear the same size regardless of distance from the camera.

## Same atlas for 2D and 3D

The `brain_atlas` class works with both ggseg and ggseg3d.
Use the same atlas object for both:

```{r}
#| label: same-atlas
#| eval: false
library(ggseg)

plot(dk)

ggseg3d(atlas = dk)
```

## Ray-traced rendering with rayshader

The interactive WebGL viewer is great for exploration, but sometimes you need a publication-quality render with realistic lighting and shadows.
`ggsegray()` takes the same atlas pipeline and outputs to rgl instead of htmlwidgets.
From there, rayshader's path tracer can produce ray-traced images directly.

```{r}
#| label: rgl-basic
#| eval: false
ggsegray(atlas = dk, hemisphere = "left") |>
  pan_camera("left lateral")
```

The same pipe functions --- `pan_camera()`, `add_glassbrain()`, `set_background()` --- work with both the htmlwidget and rgl backends.
Once the rgl scene looks right, capture it with `rgl::snapshot3d()` or hand it off to rayshader's path tracer:

```{r}
#| label: rgl-raytrace
#| eval: false
ggsegray(
  .data = some_data,
  atlas = dk,
  colour = "p",
  hemisphere = "left"
) |>
  pan_camera("left lateral") |>
  set_background("white")

rayshader::render_highquality(
  filename = "brain_raytraced.png",
  samples = 256,
  width = 1200,
  height = 900
)
rgl::close3d()
```

See `vignette("rayshader", package = "ggseg3d")` for the full walkthrough --- lighting, materials, depth of field, and camera animation.

## Finding atlases

ggseg3d ships with `dk` (cortical) and `aseg` (subcortical).
Many more are available through the [ggseg r-universe](https://ggseg.r-universe.dev):

```{r}
#| label: install-atlases
#| eval: false
install.packages("ggsegYeo2011", repos = "https://ggseg.r-universe.dev")
```

See `ggsegExtra::ggseg_atlas_repos()` for a complete list.
