---
title: "Plotting 3D brain mesh atlases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting 3D brain mesh atlases}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=F}
knitr::opts_chunk$set(eval = TRUE, fig.width = 7, fig.height = 5)
library(dplyr)
library(tidyr)
```

## Introduction

Sometimes you want to plot 3D brain surfaces from FreeSurfer. The `ggseg3d` package provides interactive WebGL-based 3D visualizations of brain atlases using Three.js via htmlwidgets.

The data uses FreeSurfer conventions (like annot and colorlut files) to create colours for the brain regions. The data structure is more complex than the 2D ggplot polygon version in `ggseg`.

A lot of credit goes to A.M. Winkler and his [Brainder](https://brainder.org/) work, which supplied us with the first examples of going from `.srf` to `.ply` files, and whose scripts massively aided us in making this work.

## Basic use

Out-of-the-box, `ggseg3d()` works without supplying any extra information. It will create a base plot of the Desikan-Killiany (dk) brain segmentations. All `[...]_3d` atlases have a built-in `colour` column for default colour plotting of the segments.

```{r start}
library(ggseg3d)
library(dplyr)
library(tidyr)
```

```{r basic, message=FALSE}
ggseg3d(show.legend = TRUE)
```

You can change the camera view using `pan_camera()`:

```{r camera-example}
ggseg3d() |>
  pan_camera("left lateral")
```

## Atlas data

The data is stored in tibbles:

```{r atlas-data}
dk_3d
```

To access all the data for a surface and hemisphere, filter to one row and then `unnest()`:

```{r unnest-example}
dk_3d |>
  filter(surf == "inflated" & hemi == "right") |>
  unnest(cols = ggseg_3d)
```

## External data supply

Notice the `mesh` column, which is a list column containing all vectors needed to create the tri-surface mesh. Also note the `label` and `region` columns, which you'll match against when providing your own data.

Be meticulous when preparing your data to ensure it matches. The function will warn you if it's struggling to match something.

The column for colour needs to be supplied to the `colour` argument, and you'll likely want to supply it to `text` as well for hover information.

```{r external-data}
someData <- dk_3d |>
  filter(surf == "inflated" & hemi == "right") |>
  unnest(ggseg_3d) |>
  ungroup() |>
  select(region) |>
  na.omit() |>
  mutate(p = sample(seq(0, .5, length.out = 100), n()) |> round(2))

ggseg3d(.data = someData, atlas = dk_3d, colour = "p", text = "p") |>
  pan_camera("right lateral")
```

## Colours

You can provide custom colour palettes either in hex or R colour names:

```{r custom-palette, eval=FALSE}
ggseg3d(
  .data = someData,
  atlas = dk_3d,
  colour = "p",
  text = "p",
  palette = c("forestgreen", "white", "firebrick")
)
```

You can also supply a named vector to control the breakpoints of the palette values, allowing the colour bar to exceed the values in the plotted data:

```{r named-palette}
ggseg3d(
  .data = someData,
  atlas = dk_3d,
  colour = "p",
  text = "p",
  palette = c("forestgreen" = 0, "white" = .05, "firebrick" = 1)
)
```

## Sub-cortical structures

When plotting sub-cortical structures, you might want to reduce the opacity of `NA` structures so you can see medial structures. You may also want to add the glass brain for reference:

```{r subcortical, warning=FALSE, message=FALSE}
somData_aseg <- aseg_3d |>
  unnest(cols = ggseg_3d) |>
  select(label) |>
  filter(!grepl("Ventricle|Putamen|Amygdala", label)) |>
  mutate(p = row_number())

ggseg3d(
  .data = somData_aseg,
  atlas = aseg_3d,
  colour = "p",
  text = "p",
  na.alpha = .5
) |>
  add_glassbrain()
```

## Camera positions

Use `pan_camera()` to set the viewing angle. Available presets:

- `left lateral`, `right lateral`  
- `left medial`, `right medial`  
- `left superior`, `right superior`  
- `left inferior`, `right inferior`  
- `left anterior`, `right anterior`  
- `left posterior`, `right posterior`  

```{r camera-presets, eval=FALSE}
ggseg3d() |> pan_camera("left lateral")
ggseg3d() |> pan_camera("right medial")
ggseg3d() |> pan_camera("left superior")
```

You can also supply custom camera positions as a list:

```{r custom-camera, eval=FALSE}
ggseg3d() |>
  pan_camera(list(eye = list(x = -200, y = 100, z = 100)))
```

## Customization

### Background colour

Change the background colour with `set_background()`:

```{r background}
ggseg3d() |>
  set_background("black")
```

### Saving images

Use `snapshot_brain()` to save static images of the widget:

```{r snapshot, eval=FALSE}
ggseg3d() |>
  pan_camera("left lateral") |>
  snapshot_brain("brain_lateral.png")
```

You can control the output size and resolution:

```{r snapshot-options, eval=FALSE}
ggseg3d() |>
  pan_camera("left lateral") |>
  snapshot_brain(
    "brain_hires.png",
    width = 1200,
    height = 1000,
    zoom = 2
  )
```

## Interactivity

The 3D plots are fully interactive:

- **Rotate**: Click and drag
- **Zoom**: Scroll wheel
- **Pan**: Right-click and drag

Hover over brain regions to see region names and any additional text data you've supplied.
