---
title: "Ray-traced brain renders with rayshader"
vignette: >
  %\VignetteIndexEntry{Ray-traced brain renders with rayshader}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
has_rgl <- rlang::is_installed("rgl")
has_rayshader <- has_rgl && rlang::is_installed("rayshader")
knitr::opts_chunk$set(
  eval = has_rgl,
  fig.width = 8,
  fig.height = 6
)
```

Screenshots from a WebGL viewer get the job done for presentations, but journals
and posters deserve better. `ggsegray()` renders brain atlases into an rgl
scene --- the same mesh data, the same colour pipeline, just a different backend.
Once the scene is in rgl, rayshader's path tracer handles realistic lighting,
soft shadows, and depth of field.

This vignette walks through the full workflow: from a basic rgl scene to a
polished ray-traced figure.

## Setup

You need rgl for the 3D scene and rayshader for the ray tracer. Both are
optional dependencies --- ggseg3d checks for them at runtime.

```{r}
#| label: load-packages
library(ggseg3d)
library(dplyr)

# install.packages(c("rgl", "rayshader")) # nolint: commented_code_linter
```

## A first render

`ggsegray()` mirrors the `ggseg3d()` API for atlas, hemisphere, and colour
mapping. The difference: instead of an htmlwidget, you get an rgl window.
Camera, background, glass brain overlays --- all of that comes through the same
pipe functions you already know from the widget side.

```{r}
#| label: first-render
#| fig-cap: "Left hemisphere of the Desikan-Killiany atlas rendered with rgl."
ggsegray(atlas = dk, hemisphere = "left") |>
  pan_camera("left lateral")
rgl::close3d()
```

This opens an interactive rgl window. Rotate with the mouse, zoom with the
scroll wheel. The camera presets match the ones from `pan_camera()` --- "left
lateral", "right medial", "left superior", and so on.

## Mapping data

Data mapping works identically to `ggseg3d()`. Provide a data frame with a
`region` column, point `colour` at a numeric or categorical variable, and
optionally set a custom palette:

```{r}
#| label: map-data
#| fig-cap: "Left hemisphere coloured by p-value."
some_data <- tibble(
  region = c("precentral", "postcentral", "insula", "superior parietal"),
  p = c(0.01, 0.04, 0.2, 0.5)
)

ggsegray(
  .data = some_data,
  atlas = dk,
  colour = "p",
  hemisphere = "left",
  palette = c("forestgreen" = 0, "white" = 0.05, "firebrick" = 1)
) |>
  pan_camera("left lateral")
rgl::close3d()
```

## Quick snapshots with rgl

Before reaching for the ray tracer, `rgl::snapshot3d()` gives you a fast screen
capture of whatever is in the rgl window. Useful for draft figures or quick
checks:

```{r}
#| label: quick-snapshot
#| fig-cap: "Quick rgl snapshot of the left hemisphere."
ggsegray(atlas = dk, hemisphere = "left") |>
  pan_camera("left lateral")

rgl::snapshot3d(tempfile(fileext = ".png"))
rgl::close3d()
```

No ray tracing, no wait --- just a screenshot. When the lighting and angle look
right, switch to `render_highquality()` for the final version.

## Ray-traced output

Build the scene with `ggsegray()` and pipe functions, then call
`render_highquality()`:

```{r}
#| label: raytrace
#| eval: !expr has_rayshader
#| fig-cap: "Ray-traced render of the left hemisphere."
ggsegray(
  .data = some_data,
  atlas = dk,
  colour = "p",
  hemisphere = "left"
) |>
  pan_camera("left lateral") |>
  set_background("white")

rayshader::render_highquality(
  filename = tempfile(fileext = ".png"),
  samples = 64,
  width = 600,
  height = 450
)

rgl::close3d()
```

Higher `samples` means less noise and longer render times. For drafts, 64--128 is
fine. For final figures, 256--512 produces clean results.

## Controlling the camera

Camera presets map to the same positions as the Three.js viewer. Pipe
`pan_camera()` after `ggsegray()`:

```{r}
#| label: camera-presets
#| fig-cap: "Camera angles: lateral, medial, superior."
ggsegray(atlas = dk, hemisphere = "left") |>
  pan_camera("left lateral")
rgl::close3d()

ggsegray(atlas = dk, hemisphere = "left") |>
  pan_camera("left medial")
rgl::close3d()

ggsegray(atlas = dk, hemisphere = "right") |>
  pan_camera("right superior")
rgl::close3d()
```

For a custom viewpoint, pass a numeric vector `c(x, y, z)`:

```{r}
#| label: camera-custom
#| fig-cap: "Left hemisphere from a custom camera position."
ggsegray(atlas = dk, hemisphere = "left") |>
  pan_camera(c(-300, 100, 150))
rgl::close3d()
```

Once the rgl window is open, `rgl::view3d()` and `rgl::observer3d()` let you
fine-tune interactively before rendering.

## Lighting

Rayshader controls lighting through `render_highquality()`. The `light_direction`
and `light_altitude` parameters set the angle and elevation of the light source:

```{r}
#| label: lighting-basic
#| eval: !expr has_rayshader
#| fig-cap: "Left hemisphere with side lighting at 90 degrees."
ggsegray(atlas = dk, hemisphere = "left") |>
  pan_camera("left lateral")

rayshader::render_highquality(
  filename = tempfile(fileext = ".png"),
  samples = 64,
  light_direction = 90,
  light_altitude = 30
)

rgl::close3d()
```

For richer lighting setups --- multiple light sources, coloured lights, area
lights --- pass custom scene elements:

```{r}
#| label: lighting-custom
#| eval: !expr has_rayshader
#| fig-cap: "Left hemisphere with a custom point light source."
ggsegray(atlas = dk, hemisphere = "left") |>
  pan_camera("left lateral")

rayshader::render_highquality(
  filename = tempfile(fileext = ".png"),
  samples = 64,
  width = 600,
  height = 450,
  light = FALSE,
  scene_elements = rayshader::rayrender::sphere(
    x = -300, y = 300, z = 200,
    radius = 50,
    material = rayshader::rayrender::light(intensity = 80, color = "white")
  ),
  interactive = FALSE
)

rgl::close3d()
```

## Background colour

White backgrounds work for most journals. For posters or slides, a dark
background can look striking. Pipe `set_background()` after `ggsegray()`:

```{r}
#| label: background-dark
#| fig-cap: "Left hemisphere with a black background."
ggsegray(atlas = dk, hemisphere = "left") |>
  pan_camera("left lateral") |>
  set_background("black")
rgl::close3d()
```

## Glass brain overlay

Subcortical structures float in space without context. A translucent glass brain
fixes that. Pipe `add_glassbrain()`:

```{r}
#| label: glassbrain
#| fig-cap: "Subcortical atlas with a translucent glass brain overlay."
ggsegray(atlas = aseg) |>
  add_glassbrain(colour = "#CCCCCC", opacity = 0.15) |>
  pan_camera("right lateral")
rgl::close3d()
```

Lower opacity values make the glass brain more see-through. For subcortical
atlases, 0.1--0.2 tends to give enough context without obscuring the structures
underneath.

## Advanced rayshader techniques

Everything below works on any rgl scene produced by `ggsegray()`. Build your
scene first, then layer on rayshader functions before rendering.

### Depth of field

Depth of field blurs regions away from a focal point, pulling attention to
a specific structure. `render_depth()` applies this as a post-processing step
on a snapshot:

```{r}
#| label: depth-of-field
#| eval: !expr has_rayshader
#| fig-cap: "Left hemisphere with depth-of-field blur."
ggsegray(
  .data = some_data,
  atlas = dk,
  colour = "p",
  hemisphere = "left"
) |>
  pan_camera("left lateral")

rayshader::render_depth(
  filename = tempfile(fileext = ".png"),
  focus = 0.5,
  focallength = 200,
  fstop = 4,
  width = 600,
  height = 450
)

rgl::close3d()
```

The `focus` parameter (0--1) controls where the focal plane sits in the depth
buffer. `focallength` and `fstop` control the strength of the blur --- lower
f-stop means shallower depth of field, just like a real camera lens.

### Adding labels

Place text labels directly in the 3D scene. These render alongside the brain
mesh and respect depth and lighting:

```{r}
#| label: labels
#| eval: !expr has_rayshader
ggsegray(atlas = dk, hemisphere = "left") |>
  pan_camera("left lateral")

rayshader::render_label(
  heightmap = matrix(1),
  text = "precentral",
  x = 150, y = 150, z = 50,
  textsize = 1.5,
  textcolor = "black",
  linecolor = "black"
)
rgl::close3d()
```

### Camera animation

`render_movie()` orbits the camera around the scene and writes frames to a
video file. Handy for conference talks or supplementary materials:

```{r}
#| label: animation
#| eval: false
ggsegray(
  .data = some_data,
  atlas = dk,
  colour = "p",
  hemisphere = "left"
) |>
  pan_camera("left lateral")

rayshader::render_movie(
  filename = "brain_rotate.mp4",
  frames = 360,
  fps = 30,
  zoom = 0.8,
  phi = 30
)

rgl::close3d()
```

This produces a 12-second rotation. Adjust `frames` and `fps` to control
duration and smoothness.

### Custom materials via rgl

Before calling rayshader, you can modify material properties on the rgl objects
directly. Every mesh added by `ggsegray()` is a standard `mesh3d` object in
the rgl scene.

```{r}
#| label: custom-materials
ggsegray(atlas = dk, hemisphere = "left") |>
  pan_camera("left lateral")

ids <- rgl::ids3d()
for (id in ids$id) {
  rgl::rgl.set.material(id = id, specular = "white", shininess = 50)
}
rgl::close3d()
```

Higher shininess gives a glossy look; set specular to black for a fully matte
surface.

### Combining views

For multi-panel figures, render each view separately and stitch them together
with magick:

```{r}
#| label: combine-views
#| eval: false
library(magick)

views <- c("left lateral", "left medial", "right lateral", "right medial")
files <- paste0("panel_", gsub(" ", "_", views), ".png")

for (i in seq_along(views)) {
  hemi <- if (grepl("left", views[i])) "left" else "right"

  ggsegray(
    .data = some_data,
    atlas = dk,
    colour = "p",
    hemisphere = hemi
  ) |>
    pan_camera(views[i]) |>
    set_background("white")

  rayshader::render_highquality(
    filename = files[i],
    samples = 256,
    width = 800,
    height = 600
  )

  rgl::close3d()
}

panels <- image_read(files)
combined <- image_montage(panels, geometry = "800x600", tile = "2x2")
image_write(combined, "figure_all_views.png")

file.remove(files)
```

## Workflow summary

A typical publication workflow:

1. **Explore** interactively with `ggseg3d()` in the htmlwidget viewer
2. **Switch** to `ggsegray()` when you have settled on the atlas, data, and
   colour scheme
3. **Draft** with `rgl::snapshot3d()` to iterate on camera angle and lighting
4. **Render** with `render_highquality()` for the final figure
5. **Combine** panels with magick if needed
