---
title: "The brain_atlas format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The brain_atlas format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Brain atlases use the `brain_atlas` class from ggseg.formats.
This format works with both ggseg (2D) and ggseg3d (3D) from a single object.

```{r}
library(ggseg3d)
library(ggseg)
library(ggplot2)
```

## One atlas, two packages

The same atlas object produces 2D polygon plots with ggseg and interactive 3D visualizations with ggseg3d:

```{r}
plot(dk)

ggseg3d(atlas = dk)
```

No need for separate objects or naming conventions.

## Atlas structure

A `brain_atlas` contains:

- **atlas**: Name of the atlas
- **type**: "cortical", "subcortical", or "tract"
- **palette**: Named colour vector for regions
- **data**: Tibble with region metadata, vertex indices, and sf geometry

```{r}
dk$atlas
dk$type
names(dk$palette)[1:5]
dk$data
```

## How 3D rendering works

For cortical atlases, 3D rendering uses vertex indices that map directly to the brain surface mesh.
Each region stores which vertices belong to it rather than a separate mesh.
This is faster and more memory-efficient.

For subcortical and tract atlases, 3D rendering uses per-region meshes since these structures don't map to a single surface.

## Atlas types

Check what type of atlas you're working with:

```{r}
dk$type

aseg$type
```

Cortical atlases have vertex indices.
Subcortical atlases have mesh data.
Tract atlases have tube meshes generated from streamlines.

## Creating atlases

Use ggsegExtra to create brain atlases from neuroimaging data:

```{r}
library(ggsegExtra)

atlas <- make_cortical_atlas(
  annot = "aparc.DKTatlas",
  include_geometry = TRUE
)

plot(atlas)

ggseg3d(atlas = atlas)
```

See the ggsegExtra documentation for:
- `make_cortical_atlas()` - from FreeSurfer annotation files
- `make_subcortical_atlas()` - from volumetric segmentation
- `make_tract_atlas()` - from tractography files
- `make_atlas_from_labels()` - from individual label files

## Benefits

**Single object**: One atlas works with both packages.
No confusion about which object to use.

**Faster 3D rendering**: Vertex indices render faster than per-region meshes for cortical atlases.

**Consistent data**: Region names, labels, and colours stay in sync between 2D and 3D.

**Simpler distribution**: Atlas packages export one object per atlas.
